Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-15 csf-2017/paper.tex
Shortening, small typesetting changes.

\begin{figure*}
  \centering
  \begin{equation*}
    \Phi  \rightarrow
    \begin{cases}
      \sem{\decode(n)}(\Phi) & \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$ and $\start \leq \addr \leq \addrend$ and $\perm \in \{ \exec,\rwx, \rwlx \}$ and $\memheap(\addr) = n$ }}\\
      \failed                                 & \text{\footnotesize{otherwise}}
    \end{cases}
  \end{equation*}\vspace{-.3cm}%
  \begin{align*}
    \stdUpdatePc{\Phi} &=
                         \begin{cases}
                           \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
                             \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$ and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}}\\
                             \failed & \text{\footnotesize{otherwise}}
                         \end{cases}
  \end{align*}
  \begin{tabular}{|c|p{4.3cm}|p{9.9cm}|}
    \hline
    $i$&$\sem{i}(\Phi)$&Conditions\\
    \hline 
    $\fail$&$\failed$&\\
    \hline
    $\halt$&$(\halted,\memheap)$&\\
    \hline
    $\move{\refreg{r_1}}{r_2}$& $\stdUpdatePc{}(\updateReg{r_1}{\memreg(r_2)})$&\\
    \hline
    $\load{\refreg{r_1}}{\refheap{r_2}}$&$\updateReg{r_1}{\memheap(\addr)}$&$\memreg(r_2) = \stdcap$\footnotesize{ and }$\perm \in \{ \rwx, \rwlx, \exec, \readwrite, \readwritel, \readonly \}$\footnotesize{ and }$\start \leq \addr \leq \addrend$\\
    \hline
    $\restricttwo{\refreg{r_1}}{r_2}$&$\stdUpdatePc{}(\updateReg{r_1}{w})$  & $\memreg(r_2) = \stdcap$\footnotesize{ and }$(\perm',g') = \decodePermPair{\memreg(r_2)}$
                                                      \footnotesize{ and }$(\perm',g') \sqsubseteq (\perm,g)$\footnotesize{ and } $w =((\perm',g'),\start,\addrend,\addr)$\\
    \hline
    $\geta{\refreg{r_1}}{\refreg{r_2}}$ & $\stdUpdatePc{\updateReg{r_1}{\addr}}$ &
                                                $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$\\
    \hline
    $\jmp{\lv}$&$\updateReg{\pcreg}{\var{newPc}}$& \footnotesize{if }$\memreg(r) = ((\perm,\gl),\start,\addrend,\addr)$\footnotesize{, then} $\var{newPc} = ((\exec,\gl),\start,\addrend,\addr)$ \footnotesize{ otherwise }$\var{newPc} = \memreg(r)$\\
    \hline
    $\store{\refheap{r_1}}{\refreg{r_2}}$&$\stdUpdatePc{\updateHeap{\addr}{\var{w}}}$&$\memreg(r_1) = \stdcap$\footnotesize{ and }$\perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}$ \footnotesize{ and } $\start \leq \addr \leq \addrend$\footnotesize{ and }$\var{w} = \memreg(r_2)$
                                                                                       \footnotesize{and if }$\var{w} = ((\_,\local),\_,\_,\_)$\footnotesize{, then  }$\perm \in \{\rwlx,\readwritel \}$\\
    \hline
    \multicolumn{3}{|c|}{$\cdots$}\\
    \hline
    \_&$\failed$&\footnotesize{otherwise}\\
    \hline
  \end{tabular}
  \caption{An excerpt from the operational semantics.}
  \label{fig:op-sem}
\end{figure*}

We define the syntax of our capability machine in
\figurename~\ref{fig:syntax}. We assume an infinite set of addresses
$\Addrs$ and define machine words as either integers or capabilities
of the form $((\perm,\gl),\var{base},\var{end},\addr)$. Such a capability
represents the authority to execute permissions $\perm$ on the memory
array $[\var{base},\var{end}]$, together with a current address $\addr$ and
a tag $\gl$ which indicates whether the capability is global or
local. The capability machine has no notion of pointers by itself, so
we refer to capabilities as pointers and capabilities
interchangeably. The available permissions and their ordering are
depicted in \figurename~\ref{fig:perm-hier}: the permissions include
null permission ($\noperm$), readonly ($\readonly$), read/write
($\readwrite$), read/execute ($\exec$) and read/write/execute ($\rwx$)
permissions. Additionally, there are three special read/write-local
($\readwritel$), read/write-local/execute ($\rwlx$) and enter
($\entry$) permissions which we will explain below.

We assume a set of register names $\RegName$. We define register files $\reg$ as
functions mapping register names to words and memories as functions that map
addresses to words. The state of the entire machine is represented as a
configuration that is either a running $\Phi \in \ExecConfs$ containing a memory
and a register file, or a failed or halted state, where the latter keeps hold
of the final state of memory. 
% We also define memory segments $\ms$:
% parts of memory represented as partial functions mapping addresses to words.
%\lau{we don't really make the distinction between memory and memory segments in the rest of the paper, so maybe we should leave this out?}

The instruction set of our machine is rather basic. Instructions $i$ include
relatively standard jump ($\jmp{}$), conditional jump ($\jnz{}{}$) and move
($\move{}{}$, copies words between registers) instructions. Also familiar are
load and store instructions for reading from and writing to memory ($\load{}{}$
and $\store{}{}$) and an arithmetic addition operators ($\plus{}{}{}$ and $\minus{}{}{}$). There are
three instructions for modifying capabilities: $\lea{}{}$ (modifies the current
address), $\restricttwo{}{}$ (modifies the permission and local/global flag) and
$\subseg{}{}{}$ (modifies the range of a capability). Importantly, these
instructions take care that the resulting capability always carries less
authority than the original (e.g. restrict will only weaken a permission).
Finally, there is an instruction $\isptr{}{}$ for testing whether a word is a
capability or a number and instructions $\getp{}{}$, $\getl{}{}$, $\getb{}{}$,
$\gete{}{}$ and $\geta{}{}$ for obtaining respectively the permissions,
local/global flag, base, end and current address of a capability.

\begin{comment}
  \begin{figure}
    \centering
    \begin{mathpar}
      \Phi \rightarrow
      \begin{cases}
        \sem{\decode(n)}(\Phi) & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\
          \;\;\text{\footnotesize{and $\start \leq \addr \leq \addrend$ and $\memheap(\addr) = n$}}\\
          \;\;\text{\footnotesize{and $\perm \in \{ \exec,\rwx, \rwlx \}$} }
        \end{array}\\
        \failed                & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \updatePcPerm{w} =
      \begin{cases}
        ((\exec,\gl),\start,\addrend,\addr) & \text{\footnotesize{if $w = ((\entry,\gl),\start,\addrend,\addr)$}}\\
        w & \text{\footnotesize{otherwise}}
      \end{cases}\\
      \and \stdUpdatePc{\Phi} =
      \begin{cases}
        \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\ 
          \;\;\text{\footnotesize{and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}}
        \end{array}\\
        \failed & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \inferrule{ } { \sem{\fail}(\Phi) = \failed } \and
      \inferrule{ } { \sem{\halt}(\Phi) = (\halted,\memheap) } \and
      \inferrule{ } { \sem{\move{\refreg{r_1}}{r_2}}(\Phi) =
        \stdUpdatePc{}(\updateReg{r_1}{\memreg(r_2)}) } \and
      \inferrule{ \memreg(r_2) = \stdcap \\
        \perm \in \{ \rwx, \rwlx, \exec, \readwrite,
        \readwritel, \readonly \}\\
        \start \leq \addr \leq \addrend } {
        \sem{\load{\refreg{r_1}}{\refheap{r_2}}}(\Phi) =
        \updateReg{r_1}{\memheap(\addr)}} \and
      \inferrule{ \memreg(r_2) = \stdcap[\permp]\\
        \permp' = \decodePermPair{\memreg(r_2)}\\
        \permp'\sqsubseteq \permp } {
        \sem{\restricttwo{\refreg{r_1}}{r_2}}(\Phi) = \\
        \stdUpdatePc{}(\updateReg{r_1}{(\permp',\start,\addrend,\addr)})
      } \and \inferrule{ \memreg(r_2) = ((\_,\_),\_,\_,\addr) } {
        \sem{\geta{\refreg{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateReg{r_1}{\addr}}} \and \inferrule{
        \var{newPc} = \updatePcPerm{\memreg(\lv)} } {
        \sem{\jmp{\lv}}(\Phi) = \updateReg{\pcreg}{\var{newPc}} } \and
      \inferrule{ \memreg(r_1) = \stdcap\\
        \start \leq \addr \leq \addrend\\
        \perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}\\
        \var{w} = \memreg(r_2) \\
        \var{w} = ((\_,\local),\_,\_,\_)\Rightarrow\perm \in
        \{\rwlx,\readwritel \} } {
        \sem{\store{\refheap{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateHeap{\addr}{\var{w}}} } \and
    \end{mathpar}
    \caption{An excerpt from the operational semantics.}
    \label{fig:op-sem}
  \end{figure}
\end{comment}
\figurename~\ref{fig:op-sem} shows an excerpt of the operational semantics containing a number of representative
instructions. Essentially, a configuration $\Phi$ will either decode and execute
the instruction at $\memreg{\pcreg}$ if it is executable and the current address
is in the valid range or fail otherwise. The table in the figure shows for a
given instruction $i$ the result of executing it in configuration $\Phi$. The
$\fail$ and $\halt$ instructions simply fail and halt. The $\move{}{}$
instruction simply modifies the register file as requested and updates the
$\pcreg$ to the next instruction using the meta-function $\stdUpdatePc{}$.

@@

The $\jmp{}$ instruction updates the program counter to a requested location,
but it is complicated by the presence of \emph{enter capabilities}, modeled
after the M-Machine's~\cite{Carter:1994:HSF:195473.195579}. Enter capabilities
cannot be used to read, write or execute and their current address and range
cannot be modified. They can only be used to jump to, but when that happens,
their permission is updated to $\exec$. This special behavior makes them
suitable for representing a kind of closures: an opaque package containing a
piece of code together with local encapsulated state. The idea is that one can
build an enter capability $c = ((\entry,\gl),\start,\addrend,\addr)$ where the
range $[\start,\addr-1]$ is used for storing the closure's local state (data or
capabilities) and the range $[\addr,\addrend]$ can be used for storing
instructions. The instructions and data are opaque to an adversary holding $c$
but when $c$ is jumped to, the data becomes available to the closure's code through
the updated version of $c$ that is now in the $\pcreg$\footnote{Note that the
  capability in the $\pcreg$-register can be used to load from.}.
% \lau{As we
%   talked about what is described here, is not how we make closures in the TR, so
%   we should consider whether we should refrain from using that word here. }

>>>>

\begin{figure*}
  \centering
  \begin{equation*}
    \Phi  \rightarrow
    \begin{cases}
      \sem{\decode(n)}(\Phi) & \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$ and $\start \leq \addr \leq \addrend$ and $\perm \in \{ \exec,\rwx, \rwlx \}$ and $\memheap(\addr) = n$ }}\\
      \failed                                 & \text{\footnotesize{otherwise}}
    \end{cases}
  \end{equation*}\vspace{-.3cm}%
  \begin{align*}
    \stdUpdatePc{\Phi} &=
                         \begin{cases}
                           \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
                             \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$ and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}}\\
                             \failed & \text{\footnotesize{otherwise}}
                         \end{cases}
  \end{align*}
  \begin{tabular}{|c|p{4.3cm}|p{9.9cm}|}
    \hline
    $i$&$\sem{i}(\Phi)$&Conditions\\
    \hline 
    $\fail$&$\failed$&\\
    \hline
    $\halt$&$(\halted,\memheap)$&\\
    \hline
    $\move{\refreg{r_1}}{r_2}$& $\stdUpdatePc{}(\updateReg{r_1}{\memreg(r_2)})$&\\
    \hline
    $\load{\refreg{r_1}}{\refheap{r_2}}$&$\updateReg{r_1}{\memheap(\addr)}$&$\memreg(r_2) = \stdcap$\footnotesize{ and }$\perm \in \{ \rwx, \rwlx, \exec, \readwrite, \readwritel, \readonly \}$\footnotesize{ and }$\start \leq \addr \leq \addrend$\\
    \hline
    $\restricttwo{\refreg{r_1}}{r_2}$&$\stdUpdatePc{}(\updateReg{r_1}{w})$  & $\memreg(r_2) = \stdcap$\footnotesize{ and }$(\perm',g') = \decodePermPair{\memreg(r_2)}$
                                                      \footnotesize{ and }$(\perm',g') \sqsubseteq (\perm,g)$\footnotesize{ and } $w =((\perm',g'),\start,\addrend,\addr)$\\
    \hline
    $\geta{\refreg{r_1}}{\refreg{r_2}}$ & $\stdUpdatePc{\updateReg{r_1}{\addr}}$ &
                                                $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$\\
    \hline
    $\jmp{\lv}$&$\updateReg{\pcreg}{\var{newPc}}$& \footnotesize{if }$\memreg(r) = ((\entry,\gl),\start,\addrend,\addr)$\footnotesize{, then} $\var{newPc} = ((\exec,\gl),\start,\addrend,\addr)$\newline
    \footnotesize{otherwise }$\var{newPc} = \memreg(r)$\\
    \hline
    $\store{\refheap{r_1}}{\refreg{r_2}}$&$\stdUpdatePc{\updateHeap{\addr}{\var{w}}}$&$\memreg(r_1) = \stdcap$\footnotesize{ and }$\perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}$ \footnotesize{ and } $\start \leq \addr \leq \addrend$\footnotesize{ and }$\var{w} = \memreg(r_2)$
                                                                                       \footnotesize{and if }$\var{w} = ((\_,\local),\_,\_,\_)$\footnotesize{, then  }$\perm \in \{\rwlx,\readwritel \}$\\
    \hline
    \multicolumn{3}{|c|}{$\cdots$}\\
    \hline
    \_&$\failed$&\footnotesize{otherwise}\\
    \hline
  \end{tabular}
  \caption{An excerpt from the operational semantics.}
  \label{fig:op-sem}
\end{figure*}

We define the syntax of our capability machine in \figurename~\ref{fig:syntax}.
We assume an infinite set of addresses $\Addrs$ and define machine words as
either integers or capabilities of the form
$((\perm,\gl),\var{base},\var{end},\addr)$. Such a capability represents the
authority to execute permissions $\perm$ on the memory range
$[\var{base},\var{end}]$, together with a current address $\addr$ and a tag
$\gl$ indicating whether the capability is global or local. There is no notion
of pointers other than capabilities, so we will use the terms interchangeably.
The available permissions and their ordering are depicted in
\figurename~\ref{fig:perm-hier}: the permissions include null permission
($\noperm$), readonly ($\readonly$), read/write ($\readwrite$), read/execute
($\exec$) and read/write/execute ($\rwx$) permissions. Additionally, there are
three special permissions: read/write-local ($\readwritel$), read/write-local/execute
($\rwlx$) and enter ($\entry$), which we will explain below.

We assume a finite set of register names $\RegName$. We define register files
$\reg$ and memories $\ms$ as functions mapping register names resp. addresses to
words. The state of the entire machine is represented as a configuration that is
either a running $\Phi \in \ExecConfs$ containing a memory and a register file,
or a failed or halted state, where the latter keeps hold of the final state of
memory.
% We also define memory segments $\ms$:
% parts of memory represented as partial functions mapping addresses to words.
%\lau{we don't really make the distinction between memory and memory segments in the rest of the paper, so maybe we should leave this out?}

The machine's instruction set is rather basic. Instructions $i$ include
relatively standard jump ($\jmp{}$), conditional jump ($\jnz{}{}$) and move
($\move{}{}$, copies words between registers) instructions. Also familiar are
load and store instructions for reading from and writing to memory ($\load{}{}$
and $\store{}{}$) and arithmetic addition operators ($\plus{}{}{}$ and
$\minus{}{}{}$, operating only on numbers). There are three instructions for
modifying capabilities: $\lea{}{}$ (modifies the current address),
$\restricttwo{}{}$ (modifies the permission and local/global flag) and
$\subseg{}{}{}$ (modifies the range of a capability). Importantly, these
instructions take care that the resulting capability always carries less
authority than the original (e.g. restrict will only weaken a permission).
Finally, the instruction $\isptr{}{}$ tests whether a word is a capability or a
number and instructions $\getp{}{}$, $\getl{}{}$, $\getb{}{}$, $\gete{}{}$ and
$\geta{}{}$ provide access to a capability's permissions, local/global flag, base,
end and current address, respectively.

\begin{comment}
  \begin{figure}
    \centering
    \begin{mathpar}
      \Phi \rightarrow
      \begin{cases}
        \sem{\decode(n)}(\Phi) & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\
          \;\;\text{\footnotesize{and $\start \leq \addr \leq \addrend$ and $\memheap(\addr) = n$}}\\
          \;\;\text{\footnotesize{and $\perm \in \{ \exec,\rwx, \rwlx \}$} }
        \end{array}\\
        \failed                & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \updatePcPerm{w} =
      \begin{cases}
        ((\exec,\gl),\start,\addrend,\addr) & \text{\footnotesize{if $w = ((\entry,\gl),\start,\addrend,\addr)$}}\\
        w & \text{\footnotesize{otherwise}}
      \end{cases}\\
      \and \stdUpdatePc{\Phi} =
      \begin{cases}
        \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
        \begin{array}[t]{l}
          \text{\footnotesize{if $\memreg(\pcreg) = \stdcap$}}\\ 
          \;\;\text{\footnotesize{and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}}
        \end{array}\\
        \failed & \text{\footnotesize{otherwise}}
      \end{cases}
      \and \inferrule{ } { \sem{\fail}(\Phi) = \failed } \and
      \inferrule{ } { \sem{\halt}(\Phi) = (\halted,\memheap) } \and
      \inferrule{ } { \sem{\move{\refreg{r_1}}{r_2}}(\Phi) =
        \stdUpdatePc{}(\updateReg{r_1}{\memreg(r_2)}) } \and
      \inferrule{ \memreg(r_2) = \stdcap \\
        \perm \in \{ \rwx, \rwlx, \exec, \readwrite,
        \readwritel, \readonly \}\\
        \start \leq \addr \leq \addrend } {
        \sem{\load{\refreg{r_1}}{\refheap{r_2}}}(\Phi) =
        \updateReg{r_1}{\memheap(\addr)}} \and
      \inferrule{ \memreg(r_2) = \stdcap[\permp]\\
        \permp' = \decodePermPair{\memreg(r_2)}\\
        \permp'\sqsubseteq \permp } {
        \sem{\restricttwo{\refreg{r_1}}{r_2}}(\Phi) = \\
        \stdUpdatePc{}(\updateReg{r_1}{(\permp',\start,\addrend,\addr)})
      } \and \inferrule{ \memreg(r_2) = ((\_,\_),\_,\_,\addr) } {
        \sem{\geta{\refreg{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateReg{r_1}{\addr}}} \and \inferrule{
        \var{newPc} = \updatePcPerm{\memreg(\lv)} } {
        \sem{\jmp{\lv}}(\Phi) = \updateReg{\pcreg}{\var{newPc}} } \and
      \inferrule{ \memreg(r_1) = \stdcap\\
        \start \leq \addr \leq \addrend\\
        \perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}\\
        \var{w} = \memreg(r_2) \\
        \var{w} = ((\_,\local),\_,\_,\_)\Rightarrow\perm \in
        \{\rwlx,\readwritel \} } {
        \sem{\store{\refheap{r_1}}{\refreg{r_2}}}(\Phi) =
        \stdUpdatePc{\updateHeap{\addr}{\var{w}}} } \and
    \end{mathpar}
    \caption{An excerpt from the operational semantics.}
    \label{fig:op-sem}
  \end{figure}
\end{comment}
\figurename~\ref{fig:op-sem} shows an excerpt of the operational semantics for a
number of representative instructions. Essentially, a configuration $\Phi$ will
either decode and execute the instruction at $\memreg{\pcreg}$ if it is
executable and the current address is in the valid range or fail otherwise. The
table in the figure shows for a given instruction $i$ the result of executing it
in configuration $\Phi$. The $\fail$ and $\halt$ instructions obviously fail and
halt respectively. The $\move{}{}$ instruction simply modifies the register file
as requested and updates the $\pcreg$ to the next instruction using the
meta-function $\stdUpdatePc{}$.

@@

The $\jmp{}$ instruction updates the program counter to a requested location,
but it is complicated by the presence of \emph{enter capabilities}, modeled
after the M-Machine's~\cite{Carter:1994:HSF:195473.195579}. Enter capabilities
cannot be used to read, write or execute and their current address and range
cannot be modified. They can only be used to jump to, but when that happens,
their permission is updated to $\exec$. This special behavior makes them
suitable for representing a kind of closures: an opaque package containing a
piece of code together with local encapsulated state. Such a package can be
built as an enter capability $c = ((\entry,\gl),\start,\addrend,\addr)$ where
the range $[\start,\addr-1]$ contains local state (data or capabilities) and
$[\addr,\addrend]$ contains instructions. The package is opaque to an adversary
holding $c$ but when $c$ is jumped to, the instructions can start executing and
have access to the local data through the updated version of $c$ that is then in
$\pcreg$\footnote{Note that the capability in the $\pcreg$-register can be used
  to load from.}.
% \lau{As we
%   talked about what is described here, is not how we make closures in the TR, so
%   we should consider whether we should refrain from using that word here. }
